\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathrm{var}}
\newcommand{\cov}{\mathrm{cov}}
\newcommand{\P}{\mathrm{P}}

# Prediction

## Measures of Regression Fit

### TSS, ESS, SSR

SW 6.4

### $R^2$

SW 6.4

## Model Selection

SW 7.5 (note: we cover substantially more details than the textbook about model selection)

### Limitations of $R^2$

SW 6.4

### Adjusted $R^2$

SW 6.4

### AIC, BIC

Alternative criteria functions:

* $AIC = 2k + n \log(SSR)$
* $BIC = k \log(n) + n \log(SSR)$

Choose model that minimizes these criteria functions.

### Cross-Validation

* Intuition: mimic the out-of-sample prediction problem

* Algorithm:

    * Split the data into J ``folds''

    * For the jth fold, do the following:
    
        1. Estimate the model using all observations *not in* the Jth fold ($\implies$ we obtain estimates $\hat{\beta}_0^j, \hat{\beta}_1^j, \ldots, \hat{\beta}_k^j$)
        
        2. Predict outcomes for observations in the Jth fold using the estimated model from part (1):
        \begin{align*}
          \tilde{Y}_{ij} = \hat{\beta}_0^j + \hat{\beta}_1^j X_{1ij} + \cdots + \hat{\beta_k^j} X_{kij}
        \end{align*}
        
        3. Compute the prediction error:
        \begin{align*}
          \tilde{U}_{ij} = Y_{ij} - \tilde{Y}_{ij}
        \end{align*}
        (this is the difference between actual outcomes for individuals in the Jth fold and their predicted outcome based on the model from part (1))
        
    * Do steps 1-3 for all $J$ folds.  This gives a prediction error $\tilde{U}_i$ for each observation in the data
    
    * compute the cross validation criteria (mean squared prediction error):
    \begin{align*}
      CV = \frac{1}{n} \sum_{i=1}^n \tilde{U}_{i}^2
    \end{align*}
    
    * choose the model that produces the smallest value of $CV$.
        
### Model Averaging

In the case where you are considering a large number of possible models, it is pretty common that a number of models will, by any of the above model selection criteria, be expected to perform very similarly when making predictions.

In this case, one strategy that usually does well in terms of making out-of-sample predictions is *model averaging*.

Suppose you have $M$ different models, and that each model can produce a predicted value for $Y_i$ --- let's call the predicted value from model $m$, $\hat{Y}_i^m$.  Model averaging would involve obtaining a new predicted value, call it $\hat{Y}_i$ by computing
\begin{align*}
  \hat{Y}_i = \frac{1}{M} \sum_{m=1}^M \hat{Y}_i^m
\end{align*}

* Usually, you would throw out models that you know predict poorly and only average together ones that perform reasonably well.

## Machine Learning

SW 14.1, 14.2, 14.6

Some extra resources on estimating Lasso and Ridge regressions in R:

* [glmnet tutorial](https://www.statology.org/lasso-regression-in-r/)

* [glmnetUtils vignette](https://cran.r-project.org/web/packages/glmnetUtils/vignettes/intro.html)

### Lasso

SW 14.3

### Ridge Regression

SW 14.4

## Binary Outcome Models

* Note: we are not necessarily so interested in prediction, but I find this a good spot to teach about binary outcome models before we conclude the course talking about causality

* In addition to referenced material below, please read all of SW Ch. 11

### Linear Probability Model

SW 11.1

### Probit and Logit

SW 11.2, 11.3

### Average Partial Effects

One of the complications with Probit and Logit is that it is not so simple to interpret the estimated parameters.  

Remember we are generally interested in partial effects, not the parameters themselves.  It just so happens that in many of the linear models that we have considered so far the $\beta$'s correspond to the partial effect --- this means that it is sometimes easy to forget that they are not what we are typically most interested in.

This is helpful framing for thinking about how to interpret the results from a Probit or Logit model.  

Let's focus on the Probit model.  In that case,
\begin{align*}
  \P(Y=1|X_1,X_2,X_3) = \Phi(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3)
\end{align*}
where $\Phi$ is the cdf of standard normal random variable.

*Continuous Case*: When $X_1$ is continuous, the partial effect of $X_1$ is given by
\begin{align*}
  \frac{\partial \P(Y=1|X_1,X_2,X_3)}{\partial X_1} = \phi(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3) \beta_1
\end{align*}
where $\phi$ is the pdf of a standard normal random variable.  This is more complicated than the partial effect in the context of a linear model.  It depends on $\phi$ (which looks complicated, but you can just use `R`'s `dnorm` command to handle that part).  More importantly, the partial effect depends on the values of $X_1,X_2,$ and $X_3$.  [As discussed above, this is likely a good thing in the context of a binary outcome model].  Thus, in order to get a partial effect, we need to put in some values for these.  If you have particular values of the covariates that you are interested in, you can definitely do that, but my general suggestion is to report the *Average Partial Effect*:
\begin{align*}
  APE &= \E\left[ \frac{\partial \P(Y=1|X_1,X_2,X_3)}{\partial X_1} \right] \\
  &= \E\left[ \phi(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3) \beta_1 \right]
\end{align*}
which you can estimate by
\begin{align*}
  \widehat{APE} &= \frac{1}{n} \sum_{i=1}^n \phi(\hat{\beta}_0 + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \hat{\beta}_3 X_3) \hat{\beta}_1
\end{align*}
which amounts to just computing the partial effect at each value of the covariates in your data and then averaging these partial effects together.  This can be a bit cumbersome to do in practice, and it is often convenient to use the `R` package `mfx` to compute these sorts of average partial effects for you.

*Discrete/Binary Case*: When $X_1$ is discrete (let's say binary, but extention to discrete is straightforward), the partial effect of $X_1$ is
\begin{align*}
  & \P(Y=1|X_1=1, X_2, X_3) - \P(Y=1|X_1=0, X_2, X_3) \\
  &\hspace{100pt} = \Phi(\beta_0 + \beta_1 + \beta_2 X_2 + \beta_3 X_3) - \Phi(\beta_0 + \beta_2 X_2 + \beta_3 X_3)
\end{align*}
Notice that $\beta_1$ does not show up in the last term.  As above, the partial effect depends on the values of $X_2$ and $X_3$ which suggests reporting an $APE$ as above (follows the same steps, just replacing the partial effect, as in the continuous case above)

* Extensions to Logit are virtually identical, just replace $\Phi$ with $\Lambda$ and $\phi$ with $\lambda$.

* Parameters from LPM, Probit, and Logit could be quite different (in fact, they are quite different by construction), but APE's are often very similar.

## Computation

$R^2$ and $\bar{R}^2$ are both reported as output from `R`'s `lm` command.

It is straightforward (and a useful excercise) to compute $TSS, ESS,$ and $SSR$ directly.  It is also straightforward to calculate $AIC$ and $BIC$ --- there are probably packages that will do this for you, but they are so easy that I suggest just calculating on your own.

The same applies to cross validation.  I suspect that there are packages available that will do this for you, but I think these are useful coding exercises and not all that difficult.  Also, if you do this yourself, it removes any kinds of "black box" issues from downloading `R` code where it may not be clear exactly what it is doing.

Computing Lasso and Ridge regressions is substantially more complicated than most other things that we have computed this semester.  The main `R` package for Lasso and Ridge regressions is the `glmnet` package.  For some reason, the syntax of the package is somewhat different from, for example, the `lm` command.  In my view, it is often easier to use the `glmnetUtils` package, which seems to be just a wrapper for `glmnet` but with functions that are analogous to `lm`.

Suppose that you have access to a training dataset called `train` and a testing dataset called `test`, you can use the `glmnetUtils` package in the following way:
```{r eval=FALSE}
library(glmnetUtils)
lasso_model <- cv.glmnet(Y ~ X1 + X2 + X3, data=train, use.model.frame=TRUE) # or whatever formula you want to use
coef(lasso_model) # if you are interested in estimated coefficients
predict(lasso_model, newdata=test) # get predictions
```

The main `R` function for estimating binary outcome models is the `glm` function (this stands for "generalized linear model").  The syntax is very similar to the syntax for the `lm` command.  For example, suppose $Y$ is binary and you are interested in estimating a logit model with regressors $X1$ and $X2$ which are contained in a `data.frame` called `data`.  In this case, you could use the following code
```{r eval=FALSE}
# glm logit
bin_reg <- glm(Y ~ X1 + X2, family=binomial(link="logit"), data=data)
summary(bin_reg)

# average partial effects
library(mfx)
bin_ape <- logitmfx(Y ~ X1 + X2, data=data, atmean=FALSE)
bin_ape

# glm probit
pro_reg <- glm(Y ~ X1 + X2, family=binomial(link="probit"), data=data)
summary(pro_reg)

# probit average partial effects
pro_ape <- probitmfx(Y ~ X1 + X2, data=data, atmean=FALSE)
pro_ape
```

## Extra Questions

1. What are some drawbacks of using $R^2$ as a model selection tool?

2. Does AIC or BIC tend to pick "more complicated" models?  What is the reason for this?

3. Suppose you are interested in predicting some outcome $Y$ and have access to covariates $X_1$, $X_2$, and $X_3$.  You estimate the following two models
\begin{align*}
  Y &= 30 + 4 X_1 - 2 X_2 - 10 X_3, \qquad R^2=0.5, AIC=3421 \\
  Y &= 9 + 2 X_1 - 3 X_2 - 2 X_3 + 2 X_1^2 + 1 X_2^2 - 4 X_3^2 + 2 X_1 X_2, \qquad R^2 = 0.75, AIC=4018
\end{align*}

    a) Which model seems to be predicting better?  Explain why.
    
    b) Using the model that is predicting better, what would be your prediction for $Y$ when $X_1=10, X_2=1, X_3=5$?  
    
4. In Lasso and Ridge regressions, it is common to "standardize" the regressors before estimating the model (e.g., the `glmnet` does this automatically for you).  What is the reason for doing this?

5. In Lasso and Ridge regressions, the penalty term lead to "shrinking" the estimated parameters in the model towards 0.  This tends to introduce bias while reducing variance.  Why can introducing bias while reducing variance potentially lead to better predictions?  Does this argument always apply or just apply in some cases?  Explain.

6.  In Lasso and Ridge regressions, the penalty term depends on the tuning parameter $\lambda$.  

    a) How is this tuning parameter often chosen in practice?  Why does it make sense to choose it in this way?
        
    b) What would happen to the estimated coefficients when $\lambda=0$?
    
    c) What would happen to the estimated coefficients as $\lambda \rightarrow \infty$?
    
7. Suppose you try to estimate a linear probability model, probit model, and logit model using the same specifications.  You notice that the estimated coefficients are substantially different from each other.  Does this mean that something has gone wrong?

## Answers to Some Extra Questions

**Answer to Question 3**

a) The first model appears to be predicting better because the AIC is lower. [Also, notice that $R^2$ is higher in the second model, but this is by construction because it includes extra terms relative to the first model which implies that it will fit at least as well in-sample as the first model, but may be suffering from over-fitting.]

b) 
    \begin{align*}
      \hat{Y} &= 30 + 4 (10) - 2 (1) - 10 (5) \\
      &= 18
    \end{align*}

**Answer to Question 6**

a) The tuning parameter is often chosen via cross validation.  It makes sense to choose it this way because this is effectively choosing a value of $\lambda$ that is making good pseudo-out-of-sample predictions.  As we will see below, if you make bad choices of $\lambda$, that could result in very poor predictions.

b) When $\lambda=0$, there would effectively be no penalty term and, therefore, the estimated parameters would coincide with the OLS estimates.

c) When $\lambda \rightarrow \infty$, the penalty term would overwhelm the term corresponding to minimizing SSR.  This would result in setting all the estimated parameters to be equal to 0.  This extreme approach is likely to lead to very poor predictions.