# Causal Inference with Observational Data

## Unconfoundedness

{{< include latex-commands.txt >}}

SW 6.8, SW Ch. 9

Unfortunately, we rarely have access to experimental data in applications or the ability to run experiments to evaluate causal effects programs/policies that we'd be interested in studying.  For example, it's hard to imagine convincing a large number of countries to randomly assign their interest rate, minimum wage, or immigration policies though these are all policies that economists would be interested in thinking about the causal effect of.

For the remainder of this chapter, we'll discuss common approaches to causal inference when the treatment is not randomly assigned.  We'll start with what is probably the most common approach: **unconfoundedness**.

**Unconfoundedness Assumption**:
\begin{align*}
  (Y(1),Y(0)) \independent D | X
\end{align*}
You can think of this as saying that, among individuals with the same covariates $X$, they have the same distributions of potential outcomes regardless of whether or not they participate in the treatment.  Note that the distribution of $X$ is still allowed to be different between the treated and untreated groups.  In other words, after you condition on covariates, there is nothing special (in terms of the distributions of potential outcomes) about the group that participates in the treatment relative to the group that doesn't participate in the treatment.

* This is potentially a strong assumption.  In order to believe this assumption, you need to believe that untreated individuals with the same characteristics can deliver, on average, the outcome that individuals in the treated group would have experienced if they had not participated in the treatment.  In math, you can write this as
    \begin{align*}
      \E[Y(0) | X, D=1] = \E[Y(0) | X, D=0]
    \end{align*}

If you are willing to believe this assumption, then you can recover the $ATT$.  There are a few different ways that you could implement this.  Probably the most common (and convenient) way is to link this condition to a regression (just like we did in the previous section on experiments).

To do this, let's continue to make the treatment effect homogeneity assumption as above.  In addition, let's assume a linear model for untreated potential outcomes
\begin{align*}
  Y(0) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + U
\end{align*}
and unconfoundedness implies that $\E[U|X_1,X_2,X_3,D] = 0$ (the conditioning on $D$ is the unconfoundedness part).  Now, recalling the definition of the observed outcome, we can write
\begin{align*}
  Y_i &= D_i Y_i(1) + (1-D_i) Y_i(0) \\
  &= D_i (Y_i(1) - Y_i(0)) + Y_i(0) \\
  &= D_i \alpha + \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + U_i
\end{align*}
which suggests running the regression of observed $Y$ on $X_1,X_2,X_3,$ and $D$ and interpreting the estimate of $\alpha$ as the causal effect of participating in the treatment.  In practice, this will be very similar to what we have done before --- so the process would not be hard, but convincing someone (or even yourself) that unconfoundedness holds will be the bigger issue here.

As a final comment, the assumption of treatment effect homogeneity is not quite so innocuous here.  It turns out that you can show that, in the presence of treatment effect heterogeneity, $\alpha$ will be equal to a weighted average of individual treatment effects, but the weights can sometimes be "strange".  There are methods that are robust to treatment effect heterogeneity (they are beyond the scope of the current class, but they are not "way" more difficult than what we are doing here).  That said, in my experience, the regression estimators (under treatment effect homogeneity) tend to deliver similar estimates to alternative estimators that are robust to treatment effect heterogeneity at least in the setup considered in this section.


## Panel Data Approaches

SW All of Ch. 10 and 13.4

In the previous section, we invoked the assumption of unconfoundedness and were in the setup where $X$ was fully observed.  But suppose instead that you thought this alternative version of unconfoundedness held
\begin{align*}
  (Y(1),Y(0)) \independent D | (X,W)
\end{align*}
where $X$ were observed random variables, but $W$ were not observed.  Following exactly the same argument as in the previous section, this would lead to a regression like
\begin{align*}
  Y_i = \alpha D_i + \beta_0 + \beta_1 X_i + \beta_2 W_i + U_i
\end{align*}
(I'm just including one $X$ and one $W$ for simplicity, but you can easily imagine the case where there are more.)  If $W$ were observed, then we could just run this regression, but since $W$ is not observed, we run into the problem of omitted variable bias (i.e., if we just ignore $W$, we won't be estimating the causal effect $\alpha$)

In this section, we'll consider the case where a researcher has access to a different type of data called **panel data**.  Panel data is data that follows the same individual (or firm, etc.) over time.  In this case, it is often helpful to index variables by time.  For example, $Y_{it}$ is the outcome for individual $i$ in time period $t$.  $X_{it}$ is the value of a regressor for individual $i$ in time period $t$ and $D_{it}$ is the value of the treatment for individual $i$ in time period $t$.  If some variable doesn't vary over time (e.g., a regressor like race), we won't use a $t$ subscript.

Panel data potentially gives us a way around the problem of not observing some variables that we would like to condition on in the model.  This is particularly likely to be the case when $W$ does not vary over time.  Let's start with there are exactly two time periods of panel data.  In that case, we can write
\begin{align*}
  Y_{it} = \alpha D_{it} + \beta_0 + \beta_1 X_{it} + \beta_2 W_i + U_{it}
\end{align*}
where we consider the case where $D$ and $X$ both change over time.  Then, defining $\Delta Y_{it} = Y_{it} - Y_{it-1}$ (and using similar notation for other variables), notice that
\begin{align*}
  \Delta Y_{it} = \alpha \Delta D_{it} + \beta_1 \Delta X_{it} + \Delta U_{it}
\end{align*}
which, importantly, no longer involves the unobserved $W_i$ and suggests running the above regression and interpreting the estimated version of $\alpha$ as an estimate of the causal effect of participating in the treatment.

* **Time fixed effects** --- The previous regression did not include an intercept.  It is common in applied work to allow for the intercept to vary over time (i.e., so that $\beta_0 = \beta_{0,t}$) which allows for "aggregate shocks" such as recessions or common trends in outcomes over time.  In practice, this amounts to just including an intercept in the previous regression, for example,

    $$
      \Delta Y_{it} = \underbrace{\theta_t} + \alpha \Delta D_{it} + \beta_1 \Delta X_{it} + \Delta U_{it}
    $$

Often, there may be many omitted, time invariant variables.  In practice, these are usually just lumped into a single **individual fixed effect** --- even if there are many time invariant, unobserved variables, we can difference them all out at the same time
\begin{align*}
  Y_{it} &= \alpha D_{it} + \beta_{0,t} + \beta_1 X_{it} + \underbrace{\beta_2 W_{1i} + \beta_3 W_{2i} + \beta_4 W_{3i}} + U_{it} \\
  &= \alpha D_{it} + \beta_{0,t} + \underbrace{\eta_i} + U_{it}
\end{align*}
and we can follow the same strategies as above.

Another case that is common in practice is when there are more than two time periods.  This case is similar to the previous one except there are multiple ways to eliminate the unobserved fixed effect.  The two most common are the

* **Within estimator**

    To motivate this approach, notice that, if, for each individual, we average their outcomes over time, the we get
    \begin{align*}
      \bar{Y}_i = \alpha \bar{D}_i + \beta_1 \bar{X}_i + (\textrm{time fixed effects}) + \bar{U}_i
    \end{align*}
    (where I have just written "time fixed effects" to indicate that these are transformed version of original fixed but still show up here.)  Subtracting this equation from the expression for $Y_{it}$ gives
    \begin{align*}
      Y_{it} - \bar{Y}_i = \alpha (D_{it} - \bar{D}_i) + \beta_1 (X_{it} - \bar{X}_i) + (\textrm{time fixed effects}) + U_{it} - \bar{U}_i
    \end{align*}

    This is a feasible regression to estimate (everything is observed here).  This is called a within estimator because the terms $\bar{Y}_i$, $\bar{D}_i$, and $\bar{X}_i$ are the within-individual averages-over-time of the corresponding variable.

* **First differences**

    Another approach to eliminating the unobserved fixed effects is to directly consider $\Delta Y_{it}$:

    \begin{align*}
      \Delta Y_{it} = \alpha \Delta D_{it} + \beta_1 \Delta X_{it} + \Delta U_{it}
    \end{align*}

    This is the same expression as we had before for the two period case.  Only here you would include observations from all available time periods on $\Delta Y_{it}, \Delta D_{it}, \Delta X_{it}$ in the regression.

It's worth mentioning the cases where a fixed effects strategy can break down:

* Unobserved variables vary over time

    $$
      Y_{it} = \alpha D_{it} + \beta_0 + \beta_1 X_{it} + \beta_2 \underbrace{W_{it}} + U_{it}
    $$
    In this case,

    $$
      \Delta Y_{it} = \alpha \Delta D_{it} + \beta_1 \Delta X_{it} + \beta_2 \underbrace{\Delta W_{it}} + \Delta U_{it}
    $$

    which still involves the unobserved $W_{it}$, and implies that the fixed effects regression will contain omitted variable bias.

* The effect of unobserved variables varies over time

    $$
      Y_{it} = \alpha D_{it} + \beta_0 + \beta_1 X_{it} + \underbrace{\beta_{2,t}} W_i + U_{it}
    $$
    In this case,

    $$
      \Delta Y_{it} = \alpha \Delta D_{it} + \beta_1 \Delta X_{it} + \underbrace{(\beta_{2,t} - \beta_{2,t-1})} W_i + \Delta U_{it}
    $$

    which still involves the unobserved $W_i$ (even though it doesn't vary over time) and, therefore, the fixed effects regressions we have been considering will contain omitted variable bias.


Also, the assumption of treatment effect homogeneity can potentially matter a lot in this context.  This will particularly be the case when (i) individuals can become treated at different points in time, and (ii) there are treatment effect dynamics (so that the effect of participating in the treatment can vary over time) --- both of these are realistic in many applications.  This is a main research area of mine and one I am happy to talk way more about.

### Difference in differences

The panel data approaches that we have been talking about so far are closely related to a **natural-experiment** type of strategy called **difference in differences** (DID).

One important difference relative to the previous approach is that DID is typically implemented when some units (these are often states or particular locations) implement a policy at some time period while others do not; and, in particular, we observe some periods before any units participate in the treatment.

Let's think about the case with exactly two time periods: $t$ and $t-1$.  In this case, we'll suppose that the outcomes that we observe are
\begin{align*}
  Y_{it} &= D_i Y_{it}(1) + (1-D_i) Y_{it}(0) \\
  Y_{it-1} &= Y_{it-1}(0)
\end{align*}
In other words, in the second period, we observe treated potential outcomes for treated units and untreated potential outcomes for untreated units (this is just like the cross-sectional case above).  But in the first period, we observe untreated potential outcomes for all units --- because no one is treated yet.

DID is often motivated by an assumption called the parallel trends assumption:

**Parallel Trends Assumption**
\begin{align*}
\E[\Delta Y_t(0) | D=1] = \E[\Delta Y_t(0) | D=0]
\end{align*}
This says that the *path* of outcomes that individuals in the treated group would have experienced if they had not been treated is the same as the path of outcomes that individual in the untreated group actually experienced.

As before, we continue to be interested in
\begin{align*}
  ATT = \E[Y_t(1) - Y_t(0) | D=1]
\end{align*}
Recall that the key identification challenge if for $\E[Y_t(0)|D=1]$ here, and notice that
\begin{align*}
  \E[Y_t(0) | D=1] &= \E[\Delta Y_t(0) | D=1] + \E[Y_{t-1}(0) | D=1] \\
  &= \E[\Delta Y_t(0) | D=0] + \E[Y_{t-1}(0)|D=1] \\
  &= \E[\Delta Y_t | D=0] + \E[Y_{t-1}|D=1]
\end{align*}
where the first equality adds and subtracts $\E[Y_{t-1}(0)|D=1]$, the second equality uses the parallel trends assumption, and the last equality holds because all the potential outcomes in the previous line are actually observed outcome.  Plugging this expression into the one for $ATT$ yields:
\begin{align*}
  ATT = \E[\Delta Y_t | D=1] - \E[\Delta Y_t | D=0]
\end{align*}
In other words, under parallel trends, the $ATT$ can be recovered by comparing the path of outcomes that treated units experienced relative to the path of outcomes that untreated units experienced (the latter of which is the path of outcomes that treated units would have experienced if they had not participated in the treatment).

As above, it is often convenient to estimate $ATT$ using a regression.  In fact, you can show that (in the case with two periods), $\alpha$ in the following regression is equal to the $ATT$:
\begin{align*}
  Y_{it} = \alpha D_{it} + \theta_t + \eta_i + v_{it}
\end{align*}
where $\E[v_t | D] = 0$.

### Computation

The lab in this chapter uses panel data to think about causal effects, so I won't provide an extended discussion here but rather just mention the syntax for a few panel data estimators.  Let's suppose that you had a data frame that, for the first few rows, looked like (this is totally made up data)

```{r echo=FALSE}
set.seed(1234)
data.frame(id=c(1,1,1,2,2,2),
           year=c(2019,2020,2021,2019,2020,2021),
           Y=(rnorm(6)*10 + 100),
           X1=(rnorm(6)*8 + 500),
           X2=sample(c(0,1),size=6,replace=TRUE)
)
```
This is what panel data typically looks like --- here, we are following a single individual (who is distinguished by their `id` variable) over three years (from 2019-2021), there is an outcome `Y` and potential regressors `X1` and `X2`.

There are several packages in `R` for estimating the fixed effects models that we have been considering.  I mainly use `plm` (for "panel linear models"), so I'll show you that one and then mention one more.

For `plm`, if you want to estimate a fixed effects model in first differences, you would use the `plm` command with the following sort of syntax

```{r eval=FALSE}
library(plm)
plm(Y ~ X1 + X2 + as.factor(year),
    data=name_of_data,
    effect="individual",
    model="fd",
    index="id")
```

We include here `as.factor(year)` to include time fixed effects, `effect="individual"` means to include an individual fixed effect, `model="fd"` says to estimate the model in first differences, and `index="id"` means that the individual identifier is in the column "id".

The code for estimating the model using a within transformation is very similar:

```{r eval=FALSE}
plm(Y ~ X1 + X2 + as.factor(year),
    data=name_of_data,
    effect="individual",
    model="within",
    index="id")
```

The only difference is that `model="fd"` has been replaced with `model="within"`.

Let me also just mention that the `estimatr` package can estimate a fixed effects model using a within transformation.  The code for this case would look like

```{r eval=FALSE}
library(estimatr)

lm_robust(Y ~ X1 + X2 + as.factor(year),
          data=name_of_data,
          fixed_effects=~id)
```

I think the advantage of using this approach is that it seems straightforward to get the heteroskedasticity robust standard errors (or cluster-robust standard errors) that are popular in economics (as we have done before for heteroskedasticity robust standard errors for a regression with just cross sectional data).  But I am not sure how (or if it is possible) to use `estimatr` to estimate the fixed effects model in first differences.



## Lab 7: Drunk Driving Laws

For this lab, we will use the `Fatalities` data.  We will study the causal effect of mandatory jail sentence policies for drunk driving on traffic fatalities.  The `Fatalities` data consists of panel data of traffic fatality death rates, whether or not a state has a mandatory jail sentence policy or not as well as several other variables from 1982-1988.  Economic theory suggests that raising the cost of some behavior (in this case, you can think of a mandatory jail sentence as raising the cost of drunk driving) will lead to less of that behavior.  That being said, it's both interesting to test this theory and also consider the magnitude of this effect.  That's what we'll do in this problem.



1. This data comes in a somewhat messier format than some of the data that we have used previously.  To start with, create a new column in the data called `afatal_per_million` that is the number of alcohol involved vehicle fatalities per millions people in a state in a particular year.  The variable `afatal` contains the total number of alcohol involved vehicle fatalities, and the variable `pop` contains the total population in a state.

2. Using a subset of the data from 1988, run a regression of `afatal_per_million` on whether or not a state has a mandatory jail sentence policy `jail`.  How do you interpret the results?

3. Using the same subset from part 2, run a regression of `afatal_per_million` on `jail`, unemployment rate (`unemp`), the tax on a case of beer (`beertax`), the percentage of southern baptists in the state (`baptist`), the percentage of residents residing in dry counties (`dry`), the percentage of young drivers in the state, (`youngdrivers`), and the average miles driven per person in a state (`miles`).  How do you interpret the estimated coefficient on `jail`?  Would you consider this to be a reasonable estimate of the (average) causal effect of mandatory jail policies on alcohol related fatalities?

4. Now, using the full data, let's estimate a fixed effects model with alcohol related fatalities per million as the outcome and mandatory jail policies as a regressor.  Estimate the model using first differences and make sure to include time fixed effects.  How do you interpret the results?

5. Estimate the same model as in part 4, but using the within estimator instead of first differences.  Compare these results to the ones from part 4.

6. Using the same within estimator as in part 5, include the same set of covariates from part 3 and interpret the estimated effect of mandatory jail policies.  How do these estimates compare to the earlier ones?

7. Now, we'll switch to using a difference in differences approach to estimating the effect of mandatory jail policies.  First, we'll manipulate the data some.

    a) To keep things simple, let's start by limiting the data to the years 1982 and 1988 and drop the in-between periods.

    b) Second, let's calculate the change in alcohol related fatalities per million between 1982 and 1998 and keep the covariates that we have been using from 1982.  One way to do this, is to use the `pivot_wider` function from the `tidyr`.  In the case of panel data, "long format" data means that each row in the data corresponds to a paricular observation *and* a particular time period.  Thus, with long format data, there are $n \times T$ total rows in the data.  On the other hand, "wide format" data means that each row holds all the data (across all time periods) for a particular observation.  Converting back and forth between long and wide formats is a common data manipulation task.  **Hint:** This step is probably unfamiliar, so I'd recommend seeing if you can use `?tidyr::pivot_wider` to see if you can figure out how to complete this step, but, if not, you can copy this code from the solutions in the next section.

    c) Finally, drop all states that are already treated in 1982.

8. Using the data that you constructed in part 7, implement the difference in differences regression of the change in alcohol related fatalities per million from 1982 to 1988 on the mandatory jail policy.  How do you interpret these results and how do they compare to the previous ones?  Now, additionally include the set of covariates that we have been using in this model.  How do you interpret these results and how do they compare to the previous ones?

9. An alternative to DID, is to include the lagged outcome as a covariate.  Using the data constructed in part 7, run a regression of alcohol related fatalities per million in 1988 on the mandatory jail policy and alcohol related fatalities per million in 1982.  How do you interpret these results and how do they compare to the previous ones?  Now include the additional covariates that we have been using in this model.  How do you interpret these results and how do they compare to the previous ones?

10. Comment on your results from parts 1-9.  Which, if any, of these are you most inclined to interpret as a reasonable estimate of the (average) causal effect of mandatory jail policies on alcohol related policies?

## Lab 7: Solutions

1.

```{r}
library(tidyr)
library(plm)

data(Fatalities, package="AER")

Fatalities$afatal_per_million <- 1000000 * (Fatalities$afatal / Fatalities$pop )
```


2.

```{r}
Fatalities88 <- subset(Fatalities, year==1988)

reg88 <- lm(afatal_per_million ~ jail, data=Fatalities88)
summary(reg88)
```
The estimated coefficient on mandatory jail laws is 9.155.  We should interpret this as just the difference between alcohol related fatalities per million in states that had mandatory jail laws in 1988 relative to states that did not have them.  We cannot reject that there is no difference between states where the policy is in place relative to those that do not have the policy.

3.

```{r}
reg88_covs <- lm(afatal_per_million ~ jail + unemp + beertax + baptist + dry + youngdrivers + miles, data=Fatalities88)
summary(reg88_covs)
```
The estimated coefficient on jail is 3.12.  It is somewhat smaller than the previous estimate, though neither is statistically significant.  We should interpret this as the partial effect of the mandatory jail policy, that is, that we estimate that mandatory jail laws increase the number of alcohol related fatalities per million by 3.12 on average controlling for the unemployment rate, beer tax, the fraction of southern baptists in the state, the fraction of residents in dry counties, the fraction of young drivers, and the average miles driven in the state.  We cannot reject that the partial effect of mandatory jail policies is equal to 0.

4.

```{r}
fd_reg <- plm(afatal_per_million ~ jail + as.factor(year),
              effect="individual",
              index="state", model="fd",
              data=Fatalities)
summary(fd_reg)
```
We should interpret the estimated coefficient on `jail` as an estimate of how much alcohol related traffic fatalities per million change on average under mandatory jail policies after accounting for time invariant variables whose effects do not change over time.  Again, we cannot reject that the effect is equal to 0.

5.

```{r}
within_reg <- plm(afatal_per_million ~ jail + as.factor(year),
              effect="individual",
              index="state", model="within",
              data=Fatalities)
summary(within_reg)
```
The estimated coefficient on `jail` has the same interpretation as in the previous problem.  The estimated effect here is marginally statistically significant.
6.

```{r}
within_reg_covs <- plm(afatal_per_million ~ jail + unemp + beertax + baptist + dry + youngdrivers + miles,
                       effect="individual",
                       index="state", model="within",
                       data=Fatalities)
summary(within_reg_covs)
```

We should interpret the estimated coefficient on `jail` as an estimate of how much alcohol related traffic fatalities per million change on average under mandatory jail policies after controlling for the unemployment rate, beer taxes, the fraction of the state that is southern baptist, the fraction of the state that lives in a dry county, the fraction of young drivers in a state, and the average number of miles driven per person in the stata, and accounting for time invariant variables whose effects do not change over time.

7.

```{r}
# part a: convert data to two period panel data
two_period <- subset(Fatalities, year==1982 | year==1988)
# and drop some missing
two_period <- subset(two_period, !is.na(jail))
two_period <- BMisc::makeBalancedPanel(two_period, "state", "year")
two_period$jail <- 1*(two_period$jail=="yes")

# part b: convert into wide format
wide_df <- pivot_wider(two_period,
                       id_cols="state",
                       names_from="year",
                       values_from=c("jail", "afatal_per_million"))

# add back other covariates from 1982
wide_df <- merge(wide_df, subset(Fatalities, year==1982)[,c("unemp", "beertax", "baptist", "dry", "youngdrivers", "miles","state")], by="state")

# change in fatal accidents over time
wide_df$Dafatal_per_million <- wide_df$afatal_per_million_1988 - wide_df$afatal_per_million_1982

# part c: drop already treated states
wide_df <- subset(wide_df, jail_1982==0)
```

8.

```{r}
did <- lm(Dafatal_per_million ~ jail_1988, data=wide_df)
summary(did)
did_covs <- lm(Dafatal_per_million ~ jail_1988 + unemp + beertax + baptist + dry + youngdrivers + miles, data=wide_df)
summary(did_covs)
```
If we are willing to believe that, in the absence of the policy, that trends in alcohol related fatalities per million people would have followed the same trends over time for treated and untreated states, then we can interpret these as causal effects.  These estimates are broadly similar to the previous ones though the second ones (that include additional covariates) are about the only ones where we ever get a negative estimate for the effect of mandatory jail policies.  Like the previous estimates, neither of these estimates are statistically different from 0.

9.

```{r}
lag_reg <- lm(afatal_per_million_1988 ~ jail_1988 + afatal_per_million_1982, data=wide_df)
summary(lag_reg)
lag_reg_covs <- lm(afatal_per_million_1988 ~ jail_1988 + afatal_per_million_1982 + unemp + beertax + baptist + dry + youngdrivers + miles, data=wide_df)
summary(lag_reg_covs)
```
These estimates directly control for alcohol related fatalities per million in the pre-treatment period 1982.  These sorts of specifications are less common in economics, but, in my view, it seems like a reasonable approach here.  That said, the results are more or less the same as earlier estimates.

10.  We don't have very strong evidence that mandatory jail policies reduced the number traffic fatalities.  In my view, probably the best specifications for trying to understand the causal effects are the ones in part 7 (particularly, the ones that include covariates there), but I think that the the results in parts 4-9 are also informative.  Broadly, these estimates are more or less similar --- none of them are statistically significant and most are positive (which is an unexpected sign).

Before we finish, let me mention a few caveats to these results:

- First, I would be very hesitant to interpret these results as definitively saying that mandatory jail policies have no effect on alcohol related traffic fatalities.  The main reason to be clear about this is that our standard error are quite large.  For example, in the second specification in part 7 (the one I like the most), a 95% confidence interval for our estimate is $[-23.1, 19.4]$.  This is a wide confidence interval --- the average number of alcohol related traffic fatalities per million across all states and time periods is only 66.  So our estimates are basically still compatible with very large reductions in alcohol related traffic fatalities up to large increases in alcohol related traffic fatalities.

- Let me make one more comment about the sign of our results.  Many of our point estimates are positive; as we discussed earlier, it is hard to rationalize harsher punishments *increasing* alcohol related traffic fatalities.  I think the main explanation for these results is just that our estimates are pretty noisy and, therefore, more or less "by chance" we are getting estimates that have an unexpected sign.  But there are some other possible explanations that are worth mentioning.  For one, there are a number of other policies related to drunk driving that occurred in the 1980s (particularly, related to legal drinking age) but perhaps others.  It is not clear how these would interact with our estimates, but they could certainly play some role.  Besides that, it seems to me that we have a pretty good set of covariates that enter our models, but there could be important covariates that we are missing.  For this reason, some expertise in how to model state-level traffic fatalities is actually a very important skill here (actually probably the key skill here!)

## Coding Questions

1. For this problem, we will use the data `rand_hie`.  This is data from the RAND health insurance experiment in the 1980s.  In the experiment, participants were randomly assigned to get Catastrophic (the least amount of coverage), insurance that came with a Deductible, insurance that came with Cost Sharing (i.e., co-insurance so that an individual pays part of their medical insurance), and Free (so that there is no cost of medical care).

    For this problem, we will be interested in whether or not changing the type of health insurance changed the amount of health care utilization and the health status of individuals.

    We will focus on the difference between the least amount of health insurance ("Catastrophic") and the most amount of health insurance ("Free").  In particular, you can start this problem by running the following code:


    `rand_hie_subset <- subset(rand_hie, plan_type %in% c("Catastrophic", "Free"))`
    `rand_hie_subset$free <- 1*(rand_hie_subset$plan_type=="Free")`

    This code creates a new data frame called `rand_hie_subset` that only contains individuals who were assigned to either the "Catastrophic" or "Free" plan types.  It also creates a new variable called `free` that is equal to one for individuals who were assigned to the "Free" plan type and zero for individuals who were assigned to the "Catastrophic" plan type.  Use this data frame for the rest of this problem.

    a) Suppose you are interested in estimating the average treatment effect on the treated of the "Free" plan relative to the "Catastrophic" plan on total medical expenditure (`total_med_expenditure`).  Can you estimate this effect by running a regression of `total_med_expenditure` on `free`?  Explain why or why not.

    b) Estimate the average treatment effect on the treated of a "Free" plan (`free`) relative to a "Catastrophic" plan on total medical expenditure (`total_med_expenditure`)  Report and interpret your results.

    c) Estimate the average treatment effect on the treated of a "Free" plan (`free`) relative to a "Catastrophic" plan on face to face doctor visits (`face_to_face_visits`)  Report and interpret your results.

    d) Estimate the average treatment effect on the treated of a "Free" plan (`free`) relative to a "Catastrophic" plan on overall health index (`health_index`)  Report and interpret your results.

    e) Provide an overall interpration of your results from parts b-d.

2. For this problem, we will study the causal effect of having more children on women's labor supply using the data `Fertility`.

    a) To start with, run a regression of the number of hours that a woman typically works per week (`work`) on whether or not she has more than two children (`morekids`).  Report your results.  Should you interpret the estimated coefficient on `morekids` as the causal effect of having more than two children?  Explain.

    b) One possible instrument in this setup is the sex composition of the first two children (i.e., whether they are both girls, both boys, or a boy and a girl).  The thinking here is that, at least in the United States, parents tend to have a preference for having both a girl and a boy and that, therefore, parents whose first two children have the same sex may be more likely to have a third child than they would have been if they have a girl and a boy.  Go through the lise of four assumptions that we discussed in the context of instrumental variables and provide some discussion about whether or not each of these assumptions are likely to hold in this context.

    c) Regardless of your answer to part (b), create a new variable called `samesex` that is equal to one for families whose first two children have the same sex.  Estimate the effect of `morekids` on `work` using `samesex` as an instrument for `morekids` and report the results.  Provide some discussion about your results.

<!--
2. For this problem, we will study the causal effect of having more children on women's labor supply using the data `Fertility`.

    a) Let's start by running a regression of the number of hours that a woman typically works per week (`work`) on whether or not she has more than two children (`morekids`), her age and $age^2$, and race/ethnicity (`afam` and `hispanic`).  Report your results.  How do you feel about interpreting the estimated coefficient on `morekids` as the causal effect of having more than two children?  Explain.

    b) One possible instrument in this setup is the sex composition of the first two children (i.e., whether they are both girls, both boys, or a boy and a girl).  The thinking here is that, at least in the United States, parents tend to have a preference for having both a girl and a boy and that, therefore, parents whose first two children have the same sex may be more likely to have a third child than they would have been if they have a girl and a boy.  Do you think that using a binary variable for whether or not the first two children have the same sex is a reasonable instrument of for `morekids` from part a?

    c) Create a new variable called `samesex` that is equal to one for families whose first two children have the same sex.  Using the same specification as in part a, use `samesex` as an instrument for `morekids` and report the results.  Provide some discussion about your results.
-->

3.  For this question, we will use the `AJR` data.  A deep question in development economics is: Why are some countries much richer than other countries?  One explanation for this is that richer countries have different institutions (e.g., property rights, democracy, etc.) that are conducive to growth.  Its hard to study these questions though because institutions do not arise randomly --- there could be reverse causality so that property rights, democracy, etc. are (perhaps partially) caused by being rich rather than the other way around.  Alternatively, other factors (say a country's geography) could cause both of these.  We'll consider one instrumental variables approach to thinking about this question in this problem.

    a) Run a regression of the log of per capita GDP (the log of per capita GDP is stored in the variable `GDP`) on a measure of the protection against expropriation risk (this is a measure of how "good" a country's institutions are (a larger number indicates "better" institutions) and it is in the variable `Exprop`).  How do you interpret these results?  Do you think it would be reasonable to interpret the estimated coefficient on `Exprop` as the causal effect of institutions on GDP.

    b) One possible instrument for `Exprop` is settler mortality (we'll use the log of this which is available in the variable `logMort`).  Settler mortality is a measure of how dangerous it was for early settlers of a particular location.  The idea is that places that have high settler mortality may have set up worse (sometimes called "extractive") institutions than places that had lower settler mortality.  But that settler mortality (from a long time ago) does not have any other direct effect on modern GDP.  Provide some discussion about whether settler mortality is a valid instrument for institutions.

    c) Estimate an IV regression of `GDP` on `Exprop` using `logMort` as an instrument for `Exprop`. How do you interpret the results?  How do these results compare to the ones from part a?


4. For this question, we'll use the data `house` to study the causal effect of incumbency on the probability that a member of the House of Representatives gets re-elected.

    a) One way to try to estimate the causal effect of incumbency is to just run a regression where the outcome is `democratic_vote_share` (this is the same outcome we'll use below) and where the model includes a binary variable for whether or not the democratic candidate is an incumbent.  What are some limitions of this strategy?

    b) The `house` data contains data about the margin of victory (is positive if they won the election and negative if they lost) for Democratic candidates in the current election and data about the Democratic margin of victory in the past election.  Explain how you could use this data in a regression discontinuity design to estimate the causal effect of incumbency.

    c) The main assumption to rationalize a regression discontinuity design is the continuity assumption.  Explain what this assumption means in the context of the regression discontinuity design that you proposed in part b.  Do you think that this assumption is likely to hold in this context?  Why or why not?

    d) Use the `house` data to implement the regression discontinuity design that you proposed in part b.  What do you estimate as the causal effect of incumbency?

5. For this problem, we will use the data `banks`.  We will study the causal effect of monetary policy on bank closures during the Great Depression.  We'll consider an interesting natural experiment in Mississippi where half the northern half of the state was in St. Louis's federal reserve district (District 8) and the southern half of the state was in Atlanta's federal reserve district (District 6).  Atlanta had much looser monetary policy (meaning they substantially increased lending) than St. Louis during the early part of the Great Depression and our interest is in whether looser monetary policy made an difference.

    a) Plot the total number of banks separately for District 6 and District 8 across all available time periods in the data.

    b) An important event in the South early in the Great Depression was the collapse of Caldwell and Company --- the largest banking chain in the South at the time.  This happened in November 1930.  The Atlanta Fed's lending markedly increased quickly after this event while St. Louis's did not.  Calculate a DID estimate of the effect of looser monetary policy on the number of banks that are still in business.  How do you interpret these results?  **Hint:** You can calculate this by taking the difference between the number of banks in District 6 relative to the number of banks in District 8 across all time periods relative to the difference between the number of banks in District 6 relative to District 8 in the first period (July 1, 1929).


## Extra Questions

1. What is the difference between treatment effect homogeneity and treatment effect heterogeneity?

2. Why do most researchers give up on trying to estimate the individual-level effect of participating in a treatment?

3. Explain what unconfoundedness means.

4. What is the key condition underlying a difference-in-differences approach to learn about the causal effect of some treatment on some outcome?

5. What are two key conditions for a valid instrument?

6. Suppose you are interested in the causal effect of participating in a union on a person's income.  Consider the following approaches.

    a) Suppose you run the following regression

        \begin{align*}
          Earnings_i = \beta_0 + \alpha Union_i + \beta_1 Education_i + U_i
        \end{align*}

        Would it be reasonable to interpret $\hat{\alpha}$ in this regression as an estimate of the causal effect of participating in a union on earnings?  Explain.

    b) Suppose you have access to panel data and run the following fixed effects regression
        \begin{align*}
          Earnings_{it} = \beta_{0,t} + \alpha Union_{it} + \beta_1 Education_{it} + \eta_i + U_{it}
        \end{align*}

        where $\eta_i$ is an individual fixed effect.  Would it be reasonable to interpert $\hat{\alpha}$ in this regression as an estimate of the causal effect of participating in a union on earnings?  Explain.  Can you think of any other advantages or disadvantages of this approach?

    c) Going back to the case with cross-sectional data, consider the regression
        \begin{align*}
          Earnings_i = \beta_0 + \alpha Union_i + U_i
        \end{align*}
        but using the variable $Z_i = 1$ if birthday is between Jan. 1 and Jun. 30 while $Z_i=0$ otherwise.  Would it be reasonable to interpert $\hat{\alpha}$ in this regression as an estimate of the causal effect of participating in a union on earnings?  Explain.  Can you think of any other advantages or disadvantages of this approach?

7. Suppose that you are interested in the effect of lower college costs on the probability of graduating from college.  You have access to student-level data from Georgia where students are eligible for the Hope Scholarship if they can keep their GPA above 3.0.

    a) What strategy can use to exploit this institional setting to learn about the causal effect of lower college costs on the probability of going to college?

    b) What sort of data would you need in order to implement this strategy?

    c) Can you think of any ways that the approach that you suggested could go wrong?

    d) Another researcher reads the results from the approach you have implemented and complains that your results are only specific to students who have grades right around the 3.0 cutoff.  Is this a fair criticism?

8. Suppose you are willing to believe versions of unconfoundedness, a linear model for untreated potential outcomes, and treatment effect homogeneity so that you could write
\begin{align*}
  Y_i = \beta_0 + \alpha D_i + \beta_1 X_i + \beta_2 W_i + U_i
\end{align*}
with $\E[U|D,X,W] = 0$ so that you were willing to interpret $\alpha$ in this regression as the causal effect of $D$ on $Y$.  However, suppose that $W$ is not observed so that you cannot operationalize the above regression.

    a) Since you do not observe $W$, you are considering just running a regression of $Y$ on $D$ and $X$ and interpreting the estimated coefficient on $D$ as the causal effect of $D$ on $Y$.  Does this seem like a good idea?

    b) In part (a), we can write a version of the model that you are thinking about estimating as
    \begin{align*}
      Y_i = \delta_0 + \delta_1 D_i + \delta_2 X_i + \epsilon_i
    \end{align*}
    Suppose that $\E[\epsilon | D, X] = 0$ and suppose also that
     \begin{align*}
       W_i = \gamma_0 + \gamma_1 D_i + \gamma_2 X_i + V_i
     \end{align*}
     with $\E[V|D,X]=0$.  Provide an expression for $\delta_1$ in terms of $\alpha$, $\gamma$'s and $\beta$'s.  Explain what this expression means.

9. Suppose you have access to an experiment where some participants were randomly assigned to participate in a job training program and others were randomly assigned not to participate.  However, some individuals that were assigned to participate in the treatment decided not to actually participate.  Let's use the following notation: $D=1$ for individuals who actually participated and $D=0$ for individuals who did not participate.  $Z=1$ for individuals who were assigned to the treatment and $Z=0$ for individuals assigned not to participate (here, $D$ and $Z$ are not exactly the same because some individuals who were assigned to the treatment did not actually participate).

    You are considering several different approaches to dealing with this issue.  Discuss which of the following are good or bad ideas:

    a) Estimating $ATT$ by $\bar{Y}_{D=1} - \bar{Y}_{D=0}$.

    b) Run the regression $Y_i = \beta_0 + \alpha D_i + U_i$ using $Z_i$ as an instrument.

10. Suppose you and a friend have conducted an experiment (things went well so that everyone complied with the treatment that they were assigned to, etc.).  You interpret the difference $\bar{Y}_{D=1} - \bar{Y}_{D=0}$ as an estimate of the $ATT$, but your friend says that you should interpret it as an estimate of the $ATE$.  In fact, according to your friend, random treatment assignment implies that $\E[Y(1)] = \E[Y(1)|D=1] = \E[Y|D=1]$ and $\E[Y(0)] = \E[Y(0)|D=0] = \E[Y|D=0]$ which implies that $ATE = \E[Y|D=1] - \E[Y|D=0]$.  Who is right?

<!--
## Answers to Some Extra Questions

**Answer to Question 4**

The key condition is the parallel trends assumption that says that, in the absence of participating in the treatment, the *path* of outcomes that individuals in the treated group is the same, on average, as the path of outcomes that individuals in the untreated group actually experienced.

**Answer to Question 9**

a) When some individuals do not comply with their treatment assignment, this approach is probably not so great.  In particular, notice that the comparison in this part of the problem is among individuals who *actually* participated in the treatment relative to those who didn't (the latter group includes both those assigned not to participate in the treatment along with those assigned to participate in the treatment, but ultimately didn't actually participate).  This suggests that this approach would generally lead to biased estimates of the $ATT$.  In the particular context of job training, you can see this would not be such a good idea if, for example, the people who were assigned to the job training program but who did not participate tended to do this because they were able to find a job before the job training program started.

b) This approach is likely to be better.  By construction, $Z$ is not correlated with $U$ (since $Z$ is randomly assigned).  $Z$ is also likely to be positively correlated with $Z$ (in particular, this will be the case if being randomly assigned to treatment increases the probability of being treated).  This implies that $Z$ is a valid instrument and should be able to deliver a reasonable estimate of the effect of participating in the treatment.

**Answer to Question 10**

While your friend's explanation is not technically wrong, it seems to me that you are more right than your friend.  There is an important issue related to external validity here.  The group of people that show up to participate in the experiment could be (and likely are) quite different from the general population.  Interpreting the results of the experiment as being an $ATE$ (in the sense of across the entire population) is therefore likely to be incorrect --- or at least would require extra assumptions and/or justifications.  Interpreting them as an $ATT$ (i.e., as the effect among those who participated in the treatment) is still perfectly reasonable though.
-->